---
title: "The effect of direct admissions programs on college enrollment"
author: "Ryan Blake"
output: html_document
---


# CLEAN
#### LOAD FILES AND PACKAGES
Note: This analysis relies on ACS data provided by IPUMS. The ACS extract exceeds the 100 MB limit for GitHub and is
not attached to the git repo. The file is stored locally and is available on request.

This chunk loads dependencies, the ACS microdata, and the ACS metadata. It then loads a sheet containing binary
variables that outline the policy timeline for each state--i.e., 1 for states/years with direct admissions and 0
otherwise. The last line loads state-level NHGIS data from 2010; this data is for DiNardo Fortin Lemieux (DFL)
reweights.

```{r, message=FALSE, warning=FALSE}
#rm(list = ls(all.names = TRUE))
library(tidyverse)
library(ipumsr)
library(fixest)
library(miceadds)

setwd("/Users/ryan/DataspellProjects/direct_admissions/")
ddi <- read_ipums_ddi("usa_00019.xml", lower_vars = TRUE)
ipums <- read_ipums_micro(ddi)

policysheet <- read.csv("/Users/ryan/DataspellProjects/direct_admissions/policysheet.csv")

nhgis <- read.csv("/Users/ryan/DataspellProjects/direct_admissions/nhgis.csv")
```


#### CALCULATE DFL REWEIGHTS
This chunk creates the variable necessary for DFL reweighting. It first cleans the NHGIS data to create new variables
that describe a states urban/rural mix, high school graduation rate, college attendance rate, youth unemployment rate,
and portion of adults serving in the military. Note that these variables are taken from states at a baseline period
prior to adoption (2010, in this case). This sheet is then merged with the policy implementation timeline.

Next, a probit regression is run to predict the likelihood that a state has adopted the policy at some point during
the observation period. These fitted values are then assigned to states that have adopted the policy; a value of 0 is
assigned to state that have not adopted the policy.

Finally, this chunk runs a series of checks to assess the validity of each chosen variable.

```{r, message = FALSE, warning=FALSE}
nhgis <- nhgis %>%
  mutate(
    #percent living in rural area
    rural_pct = rural / total_region,
    #percent w/ high school degree
    hs_complete_pct =
      (malehs+malecollege1+malecollege2+maleassoc+malebach+malemaster+maleprof+maledoc+
        femalehs+femalecollege1+femalecollege2+femaleassoc+femalebach+femalemaster+femaleprof+femaledoc)/
        total_attainment,
    #percent college go on
    college_pct =
      (malecollege1+malecollege2+maleassoc+malebach+malemaster+maleprof+maledoc+
        femalecollege1+femalecollege2+femaleassoc+femalebach+femalemaster+femaleprof+femaledoc)/
        (malehs +femalehs),
    #percent young adults <25 unemployed in labor force
    youth_unemployment =
      (male16_labor_civ_unemployed + male20_labor_civ_unemployed + male22_labor_civ_unemployed +
        female16_labor_civ_unemployed + female20_labor_civ_unemployed + female22_labor_civ_unemployed)/
        (male16_labor_civ + male20_labor_civ + male22_labor_civ +
          female16_labor_civ + female20_labor_civ + female22_labor_civ),
    #percent young adults <25 in labor force who serve in the military
    vet_pct =
      (male16_labor_vet + male20_labor_vet + male22_labor_vet +
        female16_labor_vet + female20_labor_vet + female22_labor_vet)/
        (male16_labor_total + male20_labor_total + male22_labor_total +
          female16_labor_total + female20_labor_total + female22_labor_total),
  )%>%
  select(state, statefip, rural_pct, hs_complete_pct, youth_unemployment, vet_pct, median_income)

#MERGE POLICY AND NHGIS DATA
states <- merge(policysheet, nhgis, by=c("state","statefip"))


#CALCULATE DFL
reg_nhgis <- glm(da2020 ~ rural_pct + hs_complete_pct + youth_unemployment + vet_pct +  median_income,
                 family=binomial(link='probit'), data = states)

#PRINT DFL REGRESSION RESULTS
summary(reg_nhgis)
with(summary(reg_nhgis), 1 - deviance/null.deviance) #print r squared

#FIT DFL REGRESSION RESULTS TO STATES
predict_da <- fitted(reg_nhgis)
states  <- states %>%
  mutate(pda = predict_da,
         dfl = ifelse(da2020 == 1, 1, (predict_da/(1-predict_da))))

#DFL VARIABLE CHECK
lm(rural_pct ~ da2020, data = states)
lm(rural_pct ~ da2020, data = states, weights = dfl)

lm(hs_complete_pct ~ da2020, data = states)
lm(hs_complete_pct ~ da2020, data = states, weights = dfl)

lm(youth_unemployment ~ da2020, data = states)
lm(youth_unemployment ~ da2020, data = states, weights = dfl)

lm(vet_pct ~ da2020, data = states)
lm(vet_pct ~ da2020, data = states, weights = dfl)

lm(median_income ~ da2020, data = states)
lm(median_income~ da2020, data = states, weights = dfl)


```


#### MERGE STATE DATA WITH IPUMS DATA
This chunk converts the wide state-level data to long data, and then merges with the ACS microdata. The data is then
filtered to include only 19-year-old high school graduates who have lived in the same state for more than 1 year
(a distinction which is necessary for defining who received treatment).

The outcome variable (enroll) is then created to include those who are currently attending a post-secondary institution.
A series of independent variables are also created to be used as controls.

```{r, message = FALSE, warning=FALSE}
#WIDE DATA TO LONG DATA
states_long <- states %>%
  pivot_longer(
    starts_with("da"),
    names_to = c(".value", "year"),
    names_pattern = "^(.*?)(\\d+)$",
  ) %>%
  mutate(year = as.numeric(year))

#MERGE STATE LEVEL POLICY DATA ONE TO MANY WITH ACS DATA
df <- inner_join(states_long, ipums, by = c("statefip", "year"), multiple = "all")

#FILTER TO INCLUDE ONLY HIGH SCHOOL GRADUATES
#MUTATE TO CREATE BINARY OUTCOME VARIABLE 'ENROLL' & ADD CONTROLS
df <- df %>%
        filter(educd >= 63, age == 19, school == 1 | school == 2, migrate1 == 1 | migrate1 == 2) %>%
        mutate(enroll = ifelse(educd >= 65, 1, 0), #OUTCOME VARIABLE
               female = ifelse(sex == 2, 1,0),
               hispanic = ifelse(hispand > 0,1,0),
               black = ifelse(race == 2 & hispand == 0,1,0),
               othrace = ifelse(race > 2 & hispand == 0,1,0),
               citizen = ifelse(citizen == 3, 0, 1),
               english = ifelse(speakeng == 1, 0, 1),
               insurance = ifelse(hcovany == 1, 0, 1),
               dflrewt = perwt*dfl)
```


# RESULTS

#### SUMMARY STATISTICS

This chunk prints a series of summary statistics.

```{r, message = FALSE, warning=FALSE}
df %>%
        summarize(
                n = n(),
                college_go_on = mean(enroll),
                college_go_on_sd = sd(enroll),
                female = mean(female),
                black = mean(black),
                hispanic = mean(hispanic),
                othrace = mean(othrace),
                citizen = mean(citizen),
                english = mean(english),
                insurance = mean(insurance))

df %>%
        group_by(year) %>%
        summarize(
                n = n(),
                college_go_on = mean(enroll),
                college_go_on_sd = sd(enroll),
                female = mean(female),
                black = mean(black),
                hispanic = mean(hispanic),
                othrace = mean(othrace),
                citizen = mean(citizen),
                english = mean(english),
                insurance = mean(insurance))

df %>%
        group_by(da) %>%
        summarize(
                n = n(),
                college_go_on = mean(enroll),
                college_go_on_sd = sd(enroll),
                female = mean(female),
                black = mean(black),
                hispanic = mean(hispanic),
                othrace = mean(othrace),
                citizen = mean(citizen),
                english = mean(english),
                insurance = mean(insurance))
```


#### OLS MODELS

This chunk runs two separate OLS models and prints the results (including r^2). The first model uses person level
weights supplied in the ACS data; the second model uses the DFL reweights calculated from NHGIS data. Both models
control for gender, race, ethnicity, citizenship, health insurance, English proficiency, and state/year fixed effects.
Standard errors are clustered at the state level.

```{r, message = FALSE, warning=FALSE}
#OLS - PERSON LEVEL WEIGHTS
olsmod1 <- miceadds::glm.cluster(data=df,
                                 formula=enroll ~ da + female + hispanic + black + othrace + citizen + insurance + english +
                                         factor(statefip) + factor(year),
                                 cluster="state",
                                 weights= df$perwt,
                                 family="gaussian")
summary(olsmod1$glm_res)
with(summary(olsmod1$glm_res), 1 - deviance/null.deviance) #print r squared


# OLS - DFL REWEIGHTS
olsmod2 <- miceadds::glm.cluster(data=df,
                                 formula=enroll ~ da + female + hispanic + black + othrace + citizen + insurance + english +
                                         factor(statefip) + factor(year),
                                 cluster="state",
                                 weights= df$dflrewt,
                                 family="gaussian")
summary(olsmod2$glm_res)
with(summary(olsmod2$glm_res), 1 - deviance/null.deviance) #print r squared

```


#### LOGIT MODELS

This chunk runs two separate logit models and prints the results (including r^2). The first model uses person level
weights supplied in the ACS data; the second model uses the DFL reweights calculated from NHGIS data. Both models
control for gender, race, ethnicity, citizenship, health insurance, English proficiency, and state/year fixed effects.
Standard errors are clustered at the state level.

```{r, message = FALSE, warning=FALSE}
#LOGIT - DFL REWEIGHT
logmod1 <- miceadds::glm.cluster(data=df,
                                 formula= enroll ~ da + female + hispanic + black + othrace + citizen + insurance + english +
                                         factor(statefip) + factor(year),
                                 cluster="state",
                                 weights= df$perwt,
                                 family=binomial(link= "logit"))
summary(logmod1$glm_res)
with(summary(logmod1$glm_res), 1 - deviance/null.deviance) #print r squared

#LOGIT - DFL REWEIGHT
logmod2 <- miceadds::glm.cluster(data=df,
                                 formula= enroll ~ da + female + hispanic + black + othrace + citizen + insurance + english +
                                         factor(statefip) + factor(year),
                                 cluster="state",
                                 weights= df$dflrewt,
                                 family=binomial(link= "logit"))
summary(logmod2$glm_res)
with(summary(logmod2$glm_res), 1 - deviance/null.deviance) #print r squared
```


#### EVENT STUDY

This chunk performs an event study using the feols function from the fixest package. The first call writes a new
variable to comply with the requirements of the package. The second call runs an OLS regression (note: the package does
not currently allow for binomial regression, which is more appropriate for this study). The last two calls print a
summary of the regression and plot the results.

```{r, message = FALSE, warning=FALSE}
#### EVENT STUDY
##https://lost-stats.github.io/Model_Estimation/Research_Design/event_study.html

event_study <- df %>%
        mutate(year_treated = case_when(state == "Idaho" ~ 2016,
                                        state == "South Dakota" ~ 2018,
                                        state != "Idaho" | state != "South Dakota" ~ 0))

es_dflrewt = feols(enroll ~ sunab(year_treated, year, ref.p = -1) +
        female + hispanic + black + othrace + citizen + insurance + english |
        state + year,
                   cluster = ~statefip,
                   weights = event_study$dflrewt,
                   data = event_study)

summary(es_dflrewt)
iplot(es_dflrewt,
      xlab = 'Time to treatment',
      main = 'Event study: Staggered treatment (TWFE)')
```
